---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Emma Levine"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

ANSWER HERE
Some schools have 96% FARMS, while others only have 4%. None have zero. I am pretty confident in the data because of how spread out it is and it seems like there is definitely economic segregation. 

### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

ANSWER HERE
This shows us that as FARMS pct increases (poverty), meeting or exceeding math expectations decreases. This pretty much follows the line on the graph, making this data strong. The data in the Pennsylvania plot increases slightly, but doesn't show a trend. They may look different because the PA one may be more evenly distributed economically, while the MD one has much more separate economic areas. 

### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

ANSWER HERE
The p-value is less than 0.00000000000000022. The Adjusted R-squared is 0.6734. The coefficient for FARMS Pct is -0.59511. These three numbers tell us the relationship between poverty and math scores is statistically significant, strong, and negative and about 67% of the variation in math scores is explained by poverty. The model is worth using and shows us a clear negative pattern compared to the PA model, making poverty a better predictor in MD than PA. 

### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

ANSWER HERE
The school with the largest positive residual is Fox Chapel Elementary. It is 22% white, 29% black, 20% asian and 25% hispanic. 41% of people make between $100-200k a year. Because a little less than half of the people make six figures, it would make sense that schools are over performing as they may have access to more resources such as tutoring. I would want to know if they have additional help in their school so kids can meet the expectations, such as resource rooms or teaching aides. 

The school with the largest negative residual is Germantown Elementary. It is 26% white, 27% black, 22% Hispanic and 18% Asian. 35% of people make between $100-200k a year. 48% of people have a bachelors degree, meaning that most people did not receive a college education. The per capita income is 52,000 dollars a year. This helps explain the under performance of the school as this means there is less money to be allocated to school resources. I would want to ask the principal if they offer additional help for students. 

### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

ANSWER HERE
The Maryland data supports the claim that higher FARMS percentages are linked to lower math scores, more so than in Pennsylvania. The data found in Maryland is stronger and more consistent. I would ask the school board member if they think poverty itself is the cause, or if the school's resources would change this data. Broad Acres Elementary school stands out for its high residual despite its high FARMS rate. I’d visit it first to see why this is the case.
