---
title: "HW3 WH Salaries"
output: word_document
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
url <- "https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv"

wh_salaries <- read.csv(url)

```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE
Some salaries are for detailees and some are for employees. Some salary values are 0, which may cause problems for analysis, as well as the fact that many of them have the same salary. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
summarize(wh_salaries,
          mean_salary = mean(SALARY, na.rm = TRUE),
          median_salary = median(SALARY, na.rm = TRUE))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
summarize(wh_salaries,
  mean_salary = mean(SALARY, na.rm = TRUE),
  median_salary = median(SALARY, na.rm = TRUE),
  min_salary = min(SALARY, na.rm = TRUE),
  max_salary = max(SALARY, na.rm = TRUE)
)
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE
The mean is a little bit higher than the median, meaning that only a few people earn a significantly higher salary. The outlier here is the max salary as it is much higher. The minimum being $0 shows that there is a wide range here and perhaps accounting for volunteers or special cases.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE
I would pitch a story that investigates why some people make $0 a year, and others $225,700	a year. I would want to interview someone at the white house or someone in the workforce who is privvy to the wage gap here and find out why this gap is so big. I would try to interview former white house employees about their wages and if they have any information. 
