---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Emma Levine"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
montco_inspections <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- montco_inspections |>
  summarise(
    mean_compliance_score = mean(compliance_score),
    sd_compliance_score= sd(compliance_score),
    min_compliance_score = min(compliance_score),
    max_compliance_score = max(compliance_score)
  )
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
ggplot(compliance_stats, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = proportion, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Compliance Scores of MoCo Restauraunts",
    x = "Count",
    y = "Mean",
  ) +
  theme_minimal()
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**
The standard deviation is 5.83. This means there is a wide range of compliance scores. I would be interested to see why this range is so wide and why some restaurants have high scores and others don't, but are still open. How often is the health inspector coming? Do they let certain restaurants off with a warning?
---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_daily |> 
  summarize(
    bus_mean = mean(bus),
    bus_sd = sd(bus),
    rail_mean = mean(rail),
    rail_sd = sd(rail)
  )
```       

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata_daily %>%
  mutate(
    mean_bus = mean(total),
    sd_bus = sd(total),
    z_score = (total - mean_bus) / sd_bus,
    mean_rail = mean(total),
    sd_rail= sd(total),
    z_score = (total - mean_rail)/ sd_rail
  )
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
rail_zscore <- wmata_zscores %>%
  filter(abs(z_score) > 2) %>%
  arrange(desc(abs(z_score)))

```

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(91)

sample_91 <- wmata_daily |> sample_n(91)

sample_91 |> 
  summarize(
    bus_mean = mean(bus),
    bus_sd = sd(bus),
    rail_mean = mean(rail),
    rail_sd = sd(rail)
  )


```
The values of bus mean and rail mean are relatively similar, but bus and rail standard deviation are not similar at all. This suggests that random sampling can be helpful in getting accurate data. The random sample is very similar to the population statistics, which means random sampling can be helpful in getting close to accurate data. 

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

The unusual days are New Year's Day and Christmas Day, which suggests that the holidays lead to more public transporation use. 
---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

Per capita is more useful because it gives us a better understanding of the rate based on population.

```{r}
md_car_thefts <- md_car_thefts |>
  mutate(theft_rate = (thefts_2023 / population) * 10000)
```

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
thefts_2023 |> summarize(median = median(count), total = total(count))
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

---


## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)

You would prioritize using z-score and standard deviation to see why the variety in data is so great. Using z-score would show the standard deviation and how unusual it is. This could help you to figure out why this data is spread out so much.

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

You can use z-scores to identify what pools are under performing because it shows unusual numbers, rather than using raw attendance. Using a z-score shows just how much a pool is underperforming becaue it shows the difference. Raw attendance may show that a number is lower but is within the normal variation. 

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

I would want to know of this was a random sample or not. I also would ask how they collected this dara, if it were by survey if people have to volunteer. I also would want to know the sample size and how they make sure it is the same each time. 
---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
